---
title: "Controlling the Parallel Regression Assumption across Ordinal Engines"
output: html_document
---

The parallel regression assumption states that the effect of each
predictor is the same across all cumulative thresholds. Four engines
allow users to relax this assumption in different ways:

1. **`ordinalNet::ordinalNet()`**: `parallelTerms`, `nonparallelTerms`, `parallelPenaltyFactor` arguments
2. **`ordinal::clm()`**: `nominal` formula
3. **`VGAM::vglm()`**: `parallel` argument in the family function
4. **`brms::brm()`**: `cs()` function in the model formula

This document explores each engine's parameterization directly (without
using the ordered package), compares results where they should agree,
describes their idiosyncrasies, and proposes harmonizations.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

# Setup

```{r install-ordered}
devtools::install_local("..", force = TRUE, quiet = TRUE)
```

```{r load-packages}
library(MASS)
library(VGAM)
library(ordinalNet)
library(ordinal)
```

# Data

A survey of housing conditions in Copenhagen, with an ordered factor
response `Sat` (satisfaction: Low < Medium < High) and three categorical
predictors: `Infl` (influence on management), `Type` (type of housing),
`Cont` (contact with neighbors). The `Freq` column gives cell counts.

We use only `Infl` and `Cont` as predictors to keep models tractable.
`Type` (4 levels) is omitted to avoid saturation in fully non-parallel
models.

```{r housing-data}
data("housing")

housing <- housing[housing$Type == "Tower", ]
housing$Type <- NULL
housing_long <- housing[rep(seq_len(nrow(housing)), housing$Freq), ]
housing_long$Freq <- NULL
rownames(housing_long) <- NULL

nrow(housing_long)
table(housing_long$Sat)

f <- Sat ~ Infl + Cont
```

# Models across engines

Each engine expresses parallelism through its own interface, but all four
can fit the same models. The table summarizes the syntax; the subsections
below fit each model in every engine and compare the results.

| Engine | Interface | Fully parallel | Fully non-parallel | Partial |
|---|---|---|---|---|
| `vglm` | formula + family argument | <code>parallel&nbsp;=&nbsp;TRUE</code> | <code>parallel&nbsp;=&nbsp;FALSE</code> | <code>parallel&nbsp;=&nbsp;TRUE&nbsp;~&nbsp;&lt;parallel&nbsp;terms&gt;</code> |
| `clm` | formula + `nominal` | no `nominal` | <code>nominal&nbsp;=&nbsp;~&nbsp;&lt;all&nbsp;terms&gt;</code> | <code>nominal&nbsp;=&nbsp;~&nbsp;&lt;non-parallel&nbsp;terms&gt;</code> |
| `ordinalNet` | matrix + term flags | <code>parallelTerms&nbsp;=&nbsp;TRUE</code>, <code>nonparallelTerms&nbsp;=&nbsp;FALSE</code> | <code>parallelTerms&nbsp;=&nbsp;FALSE</code>, <code>nonparallelTerms&nbsp;=&nbsp;TRUE</code> | semi-parallel (both `TRUE`) |
| `brm` | formula + `cs()` | no `cs()` | all terms in `cs()` | `cs(<non-parallel terms>)` |

The conventions differ in three ways:

- **What the formula names**: `vglm`'s `parallel` formula names the
  **parallel** terms; `clm`'s `nominal` and `brm`'s `cs()` name the
  **non-parallel** terms; `ordinalNet` names both with separate flags.
- **Sign convention**: `clm` uses the opposite sign of `vglm` and `ordinalNet`
  (latent-variable formulation); `brm` matches `vglm`. This matters only when
  comparing coefficient estimates.
- **Estimator**: `ordinalNet` is penalized, so we use `lambdaVals = 0` (no
  penalty) to compare with unpenalized `vglm` and `clm`; `brm` fits via MCMC,
  so we compare posterior means (not MLEs) and log-likelihoods (which are
  comparable).

`vglm` formula details: the `parallel` argument lives inside the family
function (`cumulative()`, `cratio()`, `sratio()`, `acat()`), and the
intercept is always category-specific (never parallel), regardless of the
formula specification. `TRUE ~ -1 + x1` (x1 parallel) is equivalent to
`FALSE ~ 1 + (all other terms)`, verified in the "`vglm`: equivalent
formula formulations" subsection below.

```{r ordinalNet-setup}
x_mat <- model.matrix(f, data = housing_long)[, -1]
y_fac <- housing_long$Sat
```

## Fully parallel model

In the fully parallel model, each predictor has a single coefficient,
shared across all thresholds. With 2 predictors and 2 thresholds, the
model has 2 intercepts + 2 = 4 free parameters. For the logit link, this
is the proportional odds model. The engines specify it as:

- `vglm`: `parallel = TRUE`
- `clm`: no `nominal` argument
- `ordinalNet`: `parallelTerms = TRUE, nonparallelTerms = FALSE`
- `brm`: no `cs()`

<details>
<summary>`vglm`: coefficient matrix and log-likelihood</summary>

```{r vglm-par}
vglm_par <- vglm(
  f, data = housing_long,
  family = VGAM::cumulative(link = "logitlink", parallel = TRUE)
)

coef(vglm_par, matrix = TRUE)
logLik(vglm_par)
```

</details>


<details>
<summary>`clm`: coefficients and log-likelihood</summary>

```{r clm-par}
clm_par <- clm(f, data = housing_long, link = "logit")

coef(clm_par)
logLik(clm_par)
```

</details>


<details>
<summary>`ordinalNet`: coefficients and log-likelihood</summary>

```{r ordinalNet-par}
set.seed(42)
on_par <- ordinalNet(
  x = x_mat, y = y_fac,
  family = "cumulative", link = "logit",
  parallelTerms = TRUE, nonparallelTerms = FALSE,
  lambdaVals = 0, nLambda = 1
)

coef(on_par)
on_par$loglik
```

</details>


<details>
<summary>`brm`: posterior means</summary>

```{r brm-par, cache=TRUE}
library(brms)

set.seed(42)
brms_par <- brm(
  Sat ~ Infl + Cont, data = housing_long,
  family = brms::cumulative("logit"),
  chains = 4, iter = 2000, warmup = 1000,
  silent = 2, refresh = 0
)

# Posterior means for non-intercept coefficients
brms::fixef(brms_par)
```

</details>


<details>
<summary>`brm`: log-likelihood</summary>

```{r brm-par-ll}
# Pointwise log-likelihood, then sum across observations
ll_mat <- brms::log_lik(brms_par)
brms_par_LL <- sum(colMeans(ll_mat))
brms_par_LL
```

</details>


All four engines should agree:

<details>
<summary>Cross-engine comparison of coefficients and log-likelihoods</summary>

```{r cross-par}
params <- c("InflMedium", "InflHigh", "ContHigh")

cf_vglm_mat <- coef(vglm_par, matrix = TRUE)
cf_clm_vec <- coef(clm_par)
cf_on_vec <- coef(on_par)
cf_brms_vec <- brms::fixef(brms_par)[params, "Estimate"]

data.frame(
  parameter = params,
  vglm = as.numeric(cf_vglm_mat[params, 1]),
  clm = as.numeric(cf_clm_vec[params]),
  ordinalNet = as.numeric(cf_on_vec[params]),
  brm = cf_brms_vec,
  check.names = FALSE
)

# Log-likelihoods
data.frame(
  engine = c("vglm", "clm", "ordinalNet", "brm"),
  LL = signif(c(
    as.numeric(logLik(vglm_par)),
    as.numeric(logLik(clm_par)),
    on_par$loglik,
    brms_par_LL
  ), 7)
)
```

</details>


`vglm` and `ordinalNet` share the same sign convention; `clm` uses the opposite
sign (latent-variable formulation). `brm` uses the same sign convention as
`vglm`. The three MLE log-likelihoods (`vglm`, `clm`, `ordinalNet`) should be
identical. `brm` reports the posterior-mean log-likelihood, which
approximates but does not exactly equal the MLE (the gap widens for
more-parameter models like the fully non-parallel model, where MCMC
mixing is harder).

## Fully non-parallel model

In the fully non-parallel model, every predictor has category-specific
coefficients. With 2 predictors and 2 thresholds, the model has
2 intercepts + 2×2 = 6 free parameters, versus 4 for the fully parallel
model. For the logit link, this model is known as the generalized ordered
logit. The engines specify it as:

- `vglm`: `parallel = FALSE`
- `clm`: `nominal = ~ Infl + Cont`
- `ordinalNet`: `parallelTerms = FALSE, nonparallelTerms = TRUE`
- `brm`: `cs(Infl) + cs(Cont)`

<details>
<summary>`vglm`: coefficient matrix and log-likelihood</summary>

```{r vglm-npar}
vglm_npar <- vglm(
  f, data = housing_long,
  family = VGAM::cumulative(link = "logitlink", parallel = FALSE)
)

coef(vglm_npar, matrix = TRUE)
logLik(vglm_npar)
```

</details>


<details>
<summary>`clm`: coefficients and log-likelihood</summary>

```{r clm-npar}
clm_npar <- clm(f, data = housing_long, link = "logit",
               nominal = ~ Infl + Cont)

coef(clm_npar)
logLik(clm_npar)
```

</details>


<details>
<summary>`ordinalNet`: coefficients and log-likelihood</summary>

```{r ordinalNet-npar}
set.seed(42)
on_npar <- ordinalNet(
  x = x_mat, y = y_fac,
  family = "cumulative", link = "logit",
  parallelTerms = FALSE, nonparallelTerms = TRUE,
  lambdaVals = 0, nLambda = 1
)

coef(on_npar)
on_npar$loglik
```

</details>


<details>
<summary>`brm`: posterior means</summary>

```{r brm-npar, cache=TRUE}
set.seed(42)
brms_npar <- brm(
  Sat ~ cs(Infl) + cs(Cont), data = housing_long,
  family = brms::cumulative("logit"),
  chains = 4, iter = 2000, warmup = 1000,
  silent = 2, refresh = 0
)

brms::fixef(brms_npar)
```

</details>


<details>
<summary>`brm`: log-likelihood</summary>

```{r brm-npar-ll}
brms_npar_LL <- sum(colMeans(brms::log_lik(brms_npar)))
brms_npar_LL
```

</details>


<details>
<summary>Coefficient structures across engines</summary>

```{r cross-npar}
# vglm fully non-parallel coefficient matrix (rows = predictors, cols = thresholds)
cf_vglm_npar_mat <- coef(vglm_npar, matrix = TRUE)
cf_vglm_npar_mat

# ordinalNet non-parallel coefficients (vector with "predictor:threshold" names)
cf_on_npar_vec <- coef(on_npar)
cf_on_npar_vec

# clm non-parallel coefficients (vector with "threshold.predictor" names, opposite sign)
cf_clm_npar_vec <- coef(clm_npar)
cf_clm_npar_vec
```

</details>


<details>
<summary>Log-likelihoods across engines</summary>

```{r cross-ll-npar}
data.frame(
  engine = c("vglm", "clm", "ordinalNet", "brm"),
  LL = signif(c(
    as.numeric(logLik(vglm_npar)),
    as.numeric(logLik(clm_npar)),
    on_npar$loglik,
    brms_npar_LL
  ), 7)
)
```

</details>


To compare `vglm` and `ordinalNet` directly (same sign convention), extract
the category-specific coefficients into a common structure:

<details>
<summary>Aligned category-specific coefficients</summary>

```{r cross-npar-aligned}
params <- c("InflMedium", "InflHigh", "ContHigh")
thr_names <- c("1", "2")

# brm fully non-parallel posterior means (category-specific)
cf_brms_npar <- brms::fixef(brms_npar)

npar_aligned <- expand.grid(
  parameter = params,
  threshold = thr_names,
  stringsAsFactors = FALSE
)

for (i in seq_len(nrow(npar_aligned))) {
  p <- npar_aligned$parameter[i]
  t <- npar_aligned$threshold[i]
  npar_aligned$vglm[i] <- cf_vglm_npar_mat[p, as.integer(t)]
  npar_aligned$ordinalNet[i] <- cf_on_npar_vec[paste0(p, ":", t)]
  npar_aligned$brm[i] <- cf_brms_npar[paste0(p, "[", t, "]"), "Estimate"]
}
npar_aligned
```

</details>


## Partial: Infl parallel, Cont non-parallel

Here Infl is parallel and Cont is non-parallel. The engines specify it
as:

- `vglm`: `parallel = TRUE ~ -1 + Infl`
- `clm`: `nominal = ~ Cont`
- `brm`: `Sat ~ Infl + cs(Cont)`
- `ordinalNet`: cannot express this directly (its flags are all-or-nothing;
  see "`ordinalNet`: semi-parallel model" below)

<details>
<summary>`vglm`: coefficient matrix and log-likelihood</summary>

```{r vglm-partial-infl}
vglm_partial_infl <- vglm(
  f, data = housing_long,
  family = VGAM::cumulative(link = "logitlink", parallel = TRUE ~ -1 + Infl)
)

coef(vglm_partial_infl, matrix = TRUE)
logLik(vglm_partial_infl)
```

</details>


<details>
<summary>`clm`: coefficients and log-likelihood</summary>

```{r clm-partial-cont}
clm_partial_cont <- clm(f, data = housing_long, link = "logit",
                        nominal = ~ Cont)

coef(clm_partial_cont)
logLik(clm_partial_cont)
```

</details>


<details>
<summary>`brm`: posterior means</summary>

```{r brm-partial-infl, cache=TRUE}
set.seed(42)
brms_partial_infl <- brm(
  Sat ~ Infl + cs(Cont), data = housing_long,
  family = brms::cumulative("logit"),
  chains = 4, iter = 2000, warmup = 1000,
  silent = 2, refresh = 0
)

brms::fixef(brms_partial_infl)
```

</details>


<details>
<summary>`brm`: log-likelihood</summary>

```{r brm-partial-infl-ll}
brms_partial_infl_LL <- sum(colMeans(brms::log_lik(brms_partial_infl)))
brms_partial_infl_LL
```

</details>


<details>
<summary>Cross-engine comparison of coefficients and log-likelihoods</summary>

```{r cross-partial-infl}
# vglm: Infl parallel, Cont non-parallel
cf_vglm_part_infl <- coef(vglm_partial_infl, matrix = TRUE)

# clm: Cont nominal (non-parallel), Infl parallel --- same model
cf_clm_part_cont <- coef(clm_partial_cont)

# brm: Infl parallel, Cont non-parallel
cf_brms_part_infl <- brms::fixef(brms_partial_infl)

# Align by parameter name
# vglm has ContHigh (non-parallel, category-specific)
# and InflMedium, InflHigh (parallel, shared across thresholds)
params_np <- c("ContHigh")
params_p <- c("InflMedium", "InflHigh")

data.frame(
  parameter = c(params_np, params_p),
  vglm_thr1 = c(
    as.numeric(cf_vglm_part_infl[params_np, 1]),
    as.numeric(cf_vglm_part_infl[params_p, 1])
  ),
  vglm_thr2 = c(
    as.numeric(cf_vglm_part_infl[params_np, 2]),
    as.numeric(cf_vglm_part_infl[params_p, 2])
  ),
  clm = c(
    as.numeric(cf_clm_part_cont[paste0("Low|Medium.", params_np)]),
    as.numeric(cf_clm_part_cont[params_p])
  ),
  brm_thr1 = c(
    cf_brms_part_infl[paste0(params_np, "[1]"), "Estimate"],
    cf_brms_part_infl[params_p, "Estimate"]
  ),
  brm_thr2 = c(
    cf_brms_part_infl[paste0(params_np, "[2]"), "Estimate"],
    cf_brms_part_infl[params_p, "Estimate"]
  ),
  check.names = FALSE
)

data.frame(
  engine = c("vglm", "clm", "brm"),
  LL = signif(c(
    as.numeric(logLik(vglm_partial_infl)),
    as.numeric(logLik(clm_partial_cont)),
    brms_partial_infl_LL
  ), 7)
)
```

</details>


These should be identical (same model, different parameterization).

## Partial: Cont parallel, Infl non-parallel

Here Cont is parallel and Infl is non-parallel. The engines specify it
as:

- `vglm`: `parallel = TRUE ~ -1 + Cont`
- `clm`: `nominal = ~ Infl`
- `brm`: `Sat ~ cs(Infl) + Cont`
- `ordinalNet`: cannot express this directly (its flags are all-or-nothing;
  see "`ordinalNet`: semi-parallel model" below)

<details>
<summary>`vglm`: coefficient matrix and log-likelihood</summary>

```{r vglm-partial-cont}
vglm_partial_cont <- vglm(
  f, data = housing_long,
  family = VGAM::cumulative(link = "logitlink", parallel = TRUE ~ -1 + Cont)
)

coef(vglm_partial_cont, matrix = TRUE)
logLik(vglm_partial_cont)
```

</details>


<details>
<summary>`clm`: coefficients and log-likelihood</summary>

```{r clm-partial-infl}
clm_partial_infl <- clm(f, data = housing_long, link = "logit",
                        nominal = ~ Infl)

coef(clm_partial_infl)
logLik(clm_partial_infl)
```

</details>


<details>
<summary>`brm`: posterior means</summary>

```{r brm-partial-cont, cache=TRUE}
set.seed(42)
brms_partial_cont <- brm(
  Sat ~ cs(Infl) + Cont, data = housing_long,
  family = brms::cumulative("logit"),
  chains = 4, iter = 2000, warmup = 1000,
  silent = 2, refresh = 0
)

brms::fixef(brms_partial_cont)
```

</details>


<details>
<summary>`brm`: log-likelihood</summary>

```{r brm-partial-cont-ll}
brms_partial_cont_LL <- sum(colMeans(brms::log_lik(brms_partial_cont)))
brms_partial_cont_LL
```

</details>


<details>
<summary>Cross-engine comparison of coefficients and log-likelihoods</summary>

```{r cross-partial-cont}
# vglm: Cont parallel, Infl non-parallel
cf_vglm_part_cont <- coef(vglm_partial_cont, matrix = TRUE)

# clm: Infl nominal (non-parallel), Cont parallel --- same model
cf_clm_part_infl <- coef(clm_partial_infl)

# brm: Cont parallel, Infl non-parallel
cf_brms_part_cont <- brms::fixef(brms_partial_cont)

# Align by parameter name (non-intercept only)
# vglm has InflMedium, InflHigh (non-parallel, category-specific)
# and ContHigh (parallel, shared across thresholds)
params_np <- c("InflMedium", "InflHigh")
params_p <- c("ContHigh")

data.frame(
  parameter = c(params_np, params_p),
  vglm_thr1 = c(
    as.numeric(cf_vglm_part_cont[params_np, 1]),
    as.numeric(cf_vglm_part_cont[params_p, 1])
  ),
  vglm_thr2 = c(
    as.numeric(cf_vglm_part_cont[params_np, 2]),
    as.numeric(cf_vglm_part_cont[params_p, 2])
  ),
  clm = c(
    as.numeric(cf_clm_part_infl[paste0("Low|Medium.", params_np)]),
    as.numeric(cf_clm_part_infl[params_p])
  ),
  brm_thr1 = c(
    cf_brms_part_cont[paste0(params_np, "[1]"), "Estimate"],
    cf_brms_part_cont[params_p, "Estimate"]
  ),
  brm_thr2 = c(
    cf_brms_part_cont[paste0(params_np, "[2]"), "Estimate"],
    cf_brms_part_cont[params_p, "Estimate"]
  ),
  check.names = FALSE
)

data.frame(
  engine = c("vglm", "clm", "brm"),
  LL = signif(c(
    as.numeric(logLik(vglm_partial_cont)),
    as.numeric(logLik(clm_partial_infl)),
    brms_partial_cont_LL
  ), 7)
)
```

</details>


These should be identical (same model, different parameterization).

## Intercept-only model: relaxing parallelism with no variable predictors

When the model contains only an intercept and no variable predictors, the
parallel regression assumption has nothing to differentiate: there are no
slope coefficients to make category-specific. Relaxing parallelism "for all
terms" therefore reduces to asking each engine what it does with its
intercepts when instructed to estimate non-parallel terms, which the engines
handle very differently.

The four engines, fit to the intercept-only model `Sat ~ 1`:

- `vglm`: `parallel = FALSE`
- `clm`: `nominal = ~ 1`
- `ordinalNet`: `parallelTerms = FALSE, nonparallelTerms = TRUE`
- `brm`: `Sat ~ cs(1)` (a degenerate request, since `cs()` has no term to wrap)

<details>
<summary>`vglm`: `parallel = FALSE` on an intercept-only model</summary>

```{r vglm-int-only}
vglm_int_par <- vglm(
  Sat ~ 1, data = housing_long,
  family = VGAM::cumulative(link = "logitlink", parallel = TRUE)
)
vglm_int_npar <- vglm(
  Sat ~ 1, data = housing_long,
  family = VGAM::cumulative(link = "logitlink", parallel = FALSE)
)

coef(vglm_int_par, matrix = TRUE)
coef(vglm_int_npar, matrix = TRUE)
all.equal(coef(vglm_int_par, matrix = TRUE), coef(vglm_int_npar, matrix = TRUE))
```

</details>

`vglm` treats the intercept as category-specific regardless of the
`parallel` setting, so `parallel = FALSE` changes nothing: the two fits have
identical thresholds and identical log-likelihoods. Relaxing parallelism is a
no-op when there are no variable predictors.

<details>
<summary>`clm`: `nominal = ~ 1` on an intercept-only model</summary>

```{r clm-int-only}
clm_int_par <- clm(Sat ~ 1, data = housing_long, link = "logit")
clm_int_npar <- clm(Sat ~ 1, data = housing_long, link = "logit",
                    nominal = ~ 1)

coef(clm_int_par)
coef(clm_int_npar)
all.equal(unname(coef(clm_int_par)), unname(coef(clm_int_npar)))
```

</details>

`clm` accepts `nominal = ~ 1` and simply re-labels the threshold intercepts
as `Threshold.(Intercept)` terms, but their estimates are numerically
identical to the plain parallel fit. As with `vglm`, there are no variable
effects to relax, so relaxing parallelism does not change the model. The
`nominal = ~ .` form, which names "all" terms, is degenerate here: with no
full-rank predictor, the resulting fit is near-singular and fails to
converge.

<details>
<summary>`ordinalNet`: intercept-only model (zero predictor columns)</summary>

```{r ordinalNet-int-only, error=TRUE}
x_mat_int <- model.matrix(~ 1, data = housing_long)[, -1]
on_int <- ordinalNet(
  x = x_mat_int, y = housing_long$Sat,
  family = "cumulative", link = "logit",
  parallelTerms = FALSE, nonparallelTerms = TRUE,
  lambdaVals = 0, nLambda = 1
)
```

</details>

`ordinalNet` fits from a predictor matrix rather than a formula, and it
requires at least one predictor column. With zero columns (the intercept-only
case), it throws `"subscript out of bounds"` before fitting, for both the
parallel and the non-parallel flag settings. So `ordinalNet` cannot express
even the simplest intercept-only model through its matrix interface, let
alone relax parallelism in one.

<details>
<summary>`brm`: intercept-only model and `cs()`</summary>

```{r brm-int-only, cache=TRUE}
brms_int <- brm(
  Sat ~ 1, data = housing_long,
  family = brms::cumulative("logit"),
  chains = 4, iter = 2000, warmup = 1000,
  silent = 2, refresh = 0
)

brms::fixef(brms_int)
```

</details>

`brm`'s `cs()` special attaches to individual predictor terms, so with an
intercept-only model there is no term to wrap and `cs()` has no effect. The
intercepts remain a single set of category-specific thresholds, and the
posterior means match the `vglm`/`clm` estimates.

In summary, relaxing parallelism for "all terms" of an intercept-only model
is vacuous for `vglm`, `clm`, and `brm` --- each leaves the intercepts (which
are always category-specific) unchanged --- while `ordinalNet` cannot fit the
model at all because its matrix interface demands at least one predictor
column.

# Engine-specific parameterizations

## `vglm`: equivalent formula formulations

`TRUE ~ -1 + Infl` (Infl parallel) should equal `FALSE ~ 1 + Cont`
(Cont non-parallel; everything else parallel).

<details>
<summary>Equivalent formula formulations: coefficients and equality checks</summary>

```{r vglm-equivalence}
vglm_partial_infl_check <- vglm(
  f, data = housing_long,
  family = VGAM::cumulative(link = "logitlink", parallel = FALSE ~ 1 + Cont)
)

vglm_partial_cont_check <- vglm(
  f, data = housing_long,
  family = VGAM::cumulative(link = "logitlink", parallel = FALSE ~ 1 + Infl)
)

# TRUE ~ -1 + Infl should equal FALSE ~ 1 + Cont
cf_a <- coef(vglm_partial_infl, matrix = TRUE)
cf_b <- coef(vglm_partial_infl_check, matrix = TRUE)

data.frame(
  parameter = rownames(cf_a),
  `TRUE~-1+Infl` = as.numeric(cf_a[, 1]),
  `FALSE~1+Cont` = as.numeric(cf_b[, 1]),
  check.names = FALSE
)

all.equal(cf_a, cf_b)

# TRUE ~ -1 + Cont should equal FALSE ~ 1 + Infl
cf_c <- coef(vglm_partial_cont, matrix = TRUE)
cf_d <- coef(vglm_partial_cont_check, matrix = TRUE)

data.frame(
  parameter = rownames(cf_c),
  `TRUE~-1+Cont` = as.numeric(cf_c[, 1]),
  `FALSE~1+Infl` = as.numeric(cf_d[, 1]),
  check.names = FALSE
)

all.equal(cf_c, cf_d)
```

</details>


## `brm`: same variable parallel and non-parallel

`brm` also lets the same variable appear both with and without `cs()`:

```r
Sat ~ Infl + cs(Infl) + Cont
```

The variable's effect decomposes into a shared (parallel) coefficient
`b` plus a category-specific deviation `bcs_j` --- a reparameterization
of the fully non-parallel model. The other engines cannot do this: `vglm`'s
`parallel` formula, `clm`'s `nominal` formula, and `ordinalNet`'s
`parallelTerms`/`nonparallelTerms` flags each assign a variable to
exactly one side of the partition.

## `ordinalNet`: semi-parallel model

`parallelTerms` and `nonparallelTerms` cannot both be `FALSE`. The
"semi-parallel" model has both `TRUE`: the elastic net penalty (scaled by
`parallelPenaltyFactor`) shrinks the nonparallel terms toward the
parallel model. At `lambda = 0`, the semi-parallel model spans the same
space as the fully non-parallel model; the parallel terms are redundant
(absorbed into nonparallel).

<details>
<summary>`ordinalNet`: semi-parallel coefficients and log-likelihood</summary>

```{r ordinalNet-semi}
set.seed(42)
on_semi <- ordinalNet(
  x = x_mat, y = y_fac,
  family = "cumulative", link = "logit",
  parallelTerms = TRUE, nonparallelTerms = TRUE,
  lambdaVals = 0, nLambda = 1
)

coef(on_semi)
on_semi$loglik
```

</details>


With higher `parallelPenaltyFactor`, parallel terms are penalized more,
so nonparallel terms should be larger (less shrunk toward zero). This
effectively biases the model **away** from the fully parallel model.

<details>
<summary>`ordinalNet`: effect of parallelPenaltyFactor on coefficients</summary>

```{r ordinalNet-pf}
set.seed(42)
on_pf1 <- ordinalNet(
  x = x_mat, y = y_fac,
  family = "cumulative", link = "logit",
  parallelTerms = TRUE, nonparallelTerms = TRUE,
  parallelPenaltyFactor = 1,
  lambdaVals = c(0.05), nLambda = 1
)

writeLines("pf=1, lambda=0.05:")
print(coef(on_pf1))

set.seed(42)
on_pf5 <- ordinalNet(
  x = x_mat, y = y_fac,
  family = "cumulative", link = "logit",
  parallelTerms = TRUE, nonparallelTerms = TRUE,
  parallelPenaltyFactor = 5,
  lambdaVals = c(0.05), nLambda = 1
)

writeLines(c("", "pf=5, lambda=0.05:"))
print(coef(on_pf5))
```

</details>


## `parallelPenaltyFactor` mechanics

### How it works

The semi-parallel model has both `parallelTerms` and `nonparallelTerms`
`TRUE`. Its elastic net penalty is constructed (from the `ordinalNet`
source) as:

```
penaltyFactorsIntercept   <- rep(0, nLev - 1)           # unpenalized
penaltyFactorsParallel    <- penaltyFactors * parallelPenaltyFactor
penaltyFactorsNonparallel <- rep(penaltyFactors, nLev - 1)
penaltyFactors <- c(penaltyFactorsIntercept,
                     penaltyFactorsParallel,
                     penaltyFactorsNonparallel)
```

For 3 levels, 3 predictors, the per-parameter penalty is:

| Parameter | Penalty |
|---|---|
| `(Intercept):1` | 0 (always unpenalized) |
| `(Intercept):2` | 0 (always unpenalized) |
| `InflMedium` (parallel) | `penaltyFactors[1] * parallelPenaltyFactor` |
| `InflHigh` (parallel) | `penaltyFactors[2] * parallelPenaltyFactor` |
| `ContHigh` (parallel) | `penaltyFactors[3] * parallelPenaltyFactor` |
| `InflMedium:1` (nonpar) | `penaltyFactors[1]` |
| `InflHigh:1` (nonpar) | `penaltyFactors[2]` |
| `ContHigh:1` (nonpar) | `penaltyFactors[3]` |
| `InflMedium:2` (nonpar) | `penaltyFactors[1]` |
| `InflHigh:2` (nonpar) | `penaltyFactors[2]` |
| `ContHigh:2` (nonpar) | `penaltyFactors[3]` |

`parallelPenaltyFactor` scales the parallel terms relative to the
nonparallel terms: higher values penalize parallel terms more (pushing
toward the fully non-parallel model), lower values penalize them less
(pushing toward the fully parallel model).

The critical constraint is that `penaltyFactors[j]` is the **same** for
both the parallel and nonparallel terms of variable `j`, so there is no
way to independently control the penalty on Cont's parallel versus its
nonparallel terms.

### Sweep: `parallelPenaltyFactor` from `0` to `Inf`

<details>
<summary>Sweep results across parallelPenaltyFactor values</summary>

```{r pf-sweep, cache=TRUE}
pf_sweep <- c(0, 0.01, 0.1, 0.5, 1, 2, 5, 10, 50, Inf)
results_pf <- data.frame()

for (pf in pf_sweep) {
  set.seed(42)
  pf_use <- ifelse(is.infinite(pf), 1e10, pf)
  fit <- tryCatch(
    ordinalNet(
      x = x_mat, y = y_fac,
      family = "cumulative", link = "logit",
      parallelTerms = TRUE, nonparallelTerms = TRUE,
      parallelPenaltyFactor = pf_use,
      lambdaVals = c(0.05), nLambda = 1,
      warn = FALSE
    ),
    error = function(e) NULL
  )
  if (!is.null(fit)) {
    cf <- coef(fit)
    results_pf <- rbind(results_pf, data.frame(
      pf = pf,
      LL = fit$loglik,
      Infl_par = abs(cf["InflMedium"]) + abs(cf["InflHigh"]),
      Infl_np = abs(cf["InflMedium:1"]) + abs(cf["InflHigh:1"]) +
        abs(cf["InflMedium:2"]) + abs(cf["InflHigh:2"]),
      Cont_par = abs(cf["ContHigh"]),
      Cont_np = abs(cf["ContHigh:1"]) + abs(cf["ContHigh:2"]),
      npar = sum(cf != 0),
      stringsAsFactors = FALSE
    ))
  }
}

results_pf$pf_label <- ifelse(
  is.infinite(results_pf$pf), "Inf",
  sprintf("%.2f", results_pf$pf)
)
print(results_pf[, c("pf_label", "LL", "Infl_par", "Infl_np",
                      "Cont_par", "Cont_np", "npar")])
```

</details>


- `pf = 0`: parallel terms unpenalized, nonparallel penalized => fully
  parallel model
- `pf = 1`: equal penalty => intermediate model
- `pf = Inf`: parallel terms zeroed out => fully non-parallel model

(Values in between interpolate: `pf = 0.01` is nearly fully parallel,
`pf >= 2` already approaches fully non-parallel.)

### Partial parallelism is impossible

Because `penaltyFactors[j]` is shared between the parallel and nonparallel
terms of each variable, `parallelPenaltyFactor` is a "global knob" for
the parallel regression assumption, not a per-variable switch: it cannot
make one variable non-parallel while another stays parallel.

Target: Infl parallel, Cont non-parallel (`clm` reference:
`LL = -404.4711` via `nominal = ~ Cont`). Attempt: set `penaltyFactors`
to penalize Cont more than Infl:

<details>
<summary>Attempted partial parallelism: coefficients and log-likelihood</summary>

```{r partial-try}
pf_partial <- c(1, 1, 10)
set.seed(42)
fit_partial_try <- ordinalNet(
  x = x_mat, y = y_fac,
  family = "cumulative", link = "logit",
  parallelTerms = TRUE, nonparallelTerms = TRUE,
  parallelPenaltyFactor = 1,
  penaltyFactors = pf_partial,
  lambdaVals = c(0.05), nLambda = 1,
  warn = FALSE
)

print(coef(fit_partial_try))
fit_partial_try$loglik
```

</details>


**Problem**: this penalizes both Cont's parallel **and** nonparallel
terms more than Infl's --- the opposite of what we want, since Cont's
nonparallel terms are the very terms we want to keep.

**Contrast with `clm`/`vglm`/`brm`**: these engines use maximum likelihood
with a design matrix that includes category-specific terms **only** for
non-parallel variables. This is a hard structural constraint in the model
itself, not a soft penalization, so per-variable partial parallelism is
native.

**Implication for ordered**: `ordinalNet` cannot express the same model
as `clm`/`vglm` when some variables are parallel and others are not. The
harmonized interface must either:

(a) Error when partial parallelism is requested for `ordinalNet`,
(b) Approximate it (poorly) via penalty tuning, or
(c) Advise users to switch to `clm` or `vglm` for partial parallelism.

### What `parallelPenaltyFactor` _is_ good for

Despite this, `parallelPenaltyFactor` serves valuable purposes:

1. **Model selection and regularized estimation**: sweeping it (or using
   cross-validation) data-adaptively estimates the overall degree of
   parallel regression assumption violation, and the semi-parallel model
   can predict better than either extreme by borrowing strength across
   categories.
2. **Variable selection**: combined with the elastic net (`alpha < 1`),
   it selects predictors while the parallelism structure determines how
   their effects are modeled.

**Practical recommendation**: for per-variable partial parallelism, use
`clm`, `vglm`, or `brm`. Use `ordinalNet` when the entire model should be
regularized, when variable selection matters, or when the degree of
parallel regression assumption violation is uncertain and should be
estimated.

### Design matrix approach for partial parallelism

A wrapper around `ordinalNet` *could* achieve per-variable partial
parallelism by constructing the design matrix by hand: shared dummy terms
for the parallel variables, and interactions with threshold indicators
for the nonparallel variables.

<details>
<summary>Constructed design matrix dimensions</summary>

```{r design-matrix-demo}
# Demonstrate the design-matrix construction (not fitted)
# Target: Infl parallel, Cont non-parallel

# Step 1: Identify groups
parallel_vars <- c("InflMedium", "InflHigh")
nonparallel_vars <- c("ContHigh")

# Step 2: Parallel design matrix (standard dummies, shared across thresholds)
X_par <- x_mat[, parallel_vars, drop = FALSE]

# Step 3: Nonparallel design matrix (interactions with threshold indicators)
K <- nlevels(y_fac)
n_thr <- K - 1
X_np_list <- list()
for (j in seq_along(nonparallel_vars)) {
  for (t in seq_len(n_thr)) {
    thr_indicator <- as.integer(as.integer(y_fac) > t)
    X_np_list[[length(X_np_list) + 1]] <- x_mat[, nonparallel_vars[j]] * thr_indicator
  }
}
X_np <- do.call(cbind, X_np_list)
colnames(X_np) <- paste0(rep(nonparallel_vars, each = n_thr), ":", seq_len(n_thr))

# Step 4: Combine
X_partial <- cbind(X_par, X_np)

data.frame(
  full_columns = ncol(x_mat),
  partial_columns = ncol(X_partial),
  parallel = paste(parallel_vars, collapse = ", "),
  nonparallel = paste(colnames(X_np), collapse = ", ")
)
```

</details>


In practice this is not native to `ordinalNet`: the wrapper would need to
bypass the internal design-matrix construction (setting both flags
`FALSE`) and pass the intercepts and the full penalty vector to the
internal optimizer or to `glmnet` directly, and the resulting fit would
forfeit elastic net regularization for the parallel terms. The approach
is mainly of theoretical interest --- it shows that `ordinalNet`'s *model
space* can express partial parallelism, but its *interface* does not
expose it.

### Parallel-to-non-parallel spectrum

<details>
<summary>`ordinalNet` spectrum: log-likelihood and information criteria</summary>

```{r spectrum, cache=TRUE}
models_spectrum <- list(
  list(name = "Parallel", pt = TRUE, nt = FALSE, pf = 1),
  list(name = "Semi (pf=0.1)", pt = TRUE, nt = TRUE, pf = 0.1),
  list(name = "Semi (pf=0.5)", pt = TRUE, nt = TRUE, pf = 0.5),
  list(name = "Semi (pf=1)", pt = TRUE, nt = TRUE, pf = 1),
  list(name = "Semi (pf=5)", pt = TRUE, nt = TRUE, pf = 5),
  list(name = "Semi (pf=10)", pt = TRUE, nt = TRUE, pf = 10),
  list(name = "Non-parallel", pt = FALSE, nt = TRUE, pf = 1)
)

spectrum_df <- data.frame(
  model = character(7), LL = NA_real_, AIC = NA_real_,
  BIC = NA_real_, npar = NA_integer_,
  stringsAsFactors = FALSE
)
for (i in seq_along(models_spectrum)) {
  m <- models_spectrum[[i]]
  set.seed(42)
  fit <- ordinalNet(
    x = x_mat, y = y_fac,
    family = "cumulative", link = "logit",
    parallelTerms = m$pt, nonparallelTerms = m$nt,
    parallelPenaltyFactor = m$pf,
    lambdaVals = c(0.05), nLambda = 1,
    warn = FALSE
  )
  npar <- sum(coef(fit) != 0)
  spectrum_df$model[i] <- m$name
  spectrum_df$LL[i] <- fit$loglik
  spectrum_df$AIC[i] <- -2 * fit$loglik + 2 * npar
  spectrum_df$BIC[i] <- -2 * fit$loglik + log(nrow(x_mat)) * npar
  spectrum_df$npar[i] <- npar
}
print(spectrum_df)
```

</details>


`clm` reference (unpenalized MLE):

<details>
<summary>`clm` reference: log-likelihoods and information criteria</summary>

```{r spectrum-clm-ref}
n <- nrow(x_mat)
data.frame(
  model = c("Parallel (clm)", "Non-parallel (clm)", "Partial Cont (clm)", "Partial Infl (clm)"),
  LL = c(
    as.numeric(logLik(clm_par)),
    as.numeric(logLik(clm_npar)),
    as.numeric(logLik(clm_partial_cont)),
    as.numeric(logLik(clm_partial_infl))
  ),
  AIC = c(
    -2 * as.numeric(logLik(clm_par)) + 2 * 5,
    -2 * as.numeric(logLik(clm_npar)) + 2 * 8,
    -2 * as.numeric(logLik(clm_partial_cont)) + 2 * 6,
    -2 * as.numeric(logLik(clm_partial_infl)) + 2 * 7
  ),
  BIC = c(
    -2 * as.numeric(logLik(clm_par)) + log(n) * 5,
    -2 * as.numeric(logLik(clm_npar)) + log(n) * 8,
    -2 * as.numeric(logLik(clm_partial_cont)) + log(n) * 6,
    -2 * as.numeric(logLik(clm_partial_infl)) + log(n) * 7
  ),
  npar = c(5, 8, 6, 7)
)
```

</details>


Note: `ordinalNet` fits use `lambda = 0.05` (regularized), so LLs are lower
than the unpenalized MLE. The semi-parallel model with `pf = 0` recovers
the fully parallel model; with `pf >= 2` it recovers the fully non-parallel
model (all parallel terms zeroed). The "sweet spot" depends on the data
and tuning.

# Threshold constraints with parallel regression

Threshold constraints and parallel regression constraints are orthogonal:
thresholds control the intercept structure, while parallelism controls
the regression coefficients. They can be combined freely. Which engines
support threshold constraints, and how?

## Available constraints

`flexible` (the default in all engines) estimates K-1 independent
thresholds. The constrained options reduce the number of free threshold
parameters:

- **equidistant / equid**: equal spacing between thresholds (2 free
  parameters: intercept + spacing, regardless of K).
- **symmetric / symm1**: thresholds symmetric around the central
  threshold (floor(K/2) + 1 free parameters).
- **symmetric2 / symm0**: symmetric about zero; central threshold is
  zero (even K) or two central thresholds are equal and opposite
  (odd K).
- **qnorm** (`vglm` only): thresholds placed at normal-distribution
  quantiles (1 free parameter: the scale).
- **sum_to_zero** (`brm` only): thresholds constrained to sum to zero
  (K-1 free parameters, but centered).

`vglm`'s threshold syntax changed at version 1.1-10: up to 1.1-9 the
argument was `thresholds` with values `"equidistant"`, `"symmetric1"`,
and `"symmetric0"`; from 1.1-10 it is `Thresh` with the abbreviated
values `"equid"`, `"symm1"`, `"symm0"`, `"free"`, and `"qnorm"`.

| Engine | Threshold argument | Options | Exposed in ordered? |
|---|---|---|---|
| `clm` | `threshold` | `flexible`, `symmetric`, `symmetric2`, `equidistant` | Yes (`threshold_structure` dial) |
| `vglm` (≤ 1.1-9) | `thresholds` (in family) | `"equidistant"`, `"symmetric1"`, `"symmetric0"` | No (wrapper uses `...`) |
| `vglm` (≥ 1.1-10) | `Thresh` (in family) | `NULL`, `"equid"`, `"symm1"`, `"symm0"`, `"free"`, `"qnorm"` | No (wrapper uses `...`) |
| `brm` | `threshold` (in family) | `"flexible"`, `"equidistant"`, `"sum_to_zero"` | No (not registered) |
| `ordinalNet` | --- | None | --- |

## Interaction with parallelism

Because thresholds and parallelism constrain different parts of the
model, they compose without conflict:

```r
# clm: equidistant thresholds + partial non-parallel effects
clm(Sat ~ Infl + Cont, data = data,
    nominal = ~ Cont, threshold = "equidistant")

# vglm: equidistant thresholds + non-parallel effects
vglm(Sat ~ Infl + Cont, data = data,
     family = VGAM::cumulative(link = "logitlink",
                         parallel = FALSE, Thresh = "equid"))

# brm: equidistant thresholds + category-specific effects
brm(Sat ~ Infl + cs(Cont), data = data,
    family = brms::cumulative("logit", threshold = "equidistant"))
```

The parallel regression specification determines how many coefficient
vectors are estimated (one for a fully parallel model, K-1 for a fully
non-parallel model, variable for partial parallelism).
The threshold constraint determines how many intercepts are estimated.
Together, they set the total number of free parameters.

## Comparison on housing data

<details>
<summary>`clm`: threshold sweep log-likelihoods and degrees of freedom</summary>

```{r threshold-comparison, cache=TRUE}
# clm: threshold sweep with fully parallel model
clm_thr_par <- data.frame(
  threshold = c("flexible", "symmetric", "equidistant"),
  LL = NA_real_, df = NA_integer_,
  stringsAsFactors = FALSE
)
for (i in seq_along(clm_thr_par$threshold)) {
  fit <- clm(f, data = housing_long, link = "logit",
             threshold = clm_thr_par$threshold[i])
  clm_thr_par$LL[i] <- as.numeric(logLik(fit))
  clm_thr_par$df[i] <- attr(logLik(fit), "df")
}
print(clm_thr_par)

# clm: threshold sweep with fully non-parallel model (all nominal)
clm_thr_npar <- clm_thr_par
for (i in seq_along(clm_thr_npar$threshold)) {
  fit <- clm(f, data = housing_long, link = "logit",
             nominal = ~ Infl + Cont,
             threshold = clm_thr_npar$threshold[i])
  clm_thr_npar$LL[i] <- as.numeric(logLik(fit))
  clm_thr_npar$df[i] <- attr(logLik(fit), "df")
}
print(clm_thr_npar)
```

</details>


With 3 satisfaction levels (2 thresholds) and 2 categorical predictors,
the equidistant constraint saves at most 1 degree of freedom (2
thresholds down to 1 spacing parameter). The gain grows with the number
of ordinal levels. Threshold constraints are most useful when the
spacing between categories is plausibly uniform, and they interact
naturally with the parallel regression specification: a user can
simultaneously request equidistant thresholds and partial non-parallel
effects, getting a parsimonious model on both dimensions.

# Harmonization proposals

## Proposal 1: Unified parallelism specification

A single engine argument called `parallelism` that accepts:

- `parallelism = "parallel"` (or `TRUE`): all terms parallel.
  Equivalent to `vglm: parallel = TRUE`, `clm: nominal = NULL`,
  `brm: Sat ~ x1 + x2` (no `cs()` terms),
  `ordinalNet: parallelTerms=TRUE, nonparallelTerms=FALSE`.

- `parallelism = "nonparallel"` (or `FALSE`): no terms parallel.
  Equivalent to `vglm: parallel = FALSE`, `clm: nominal = ~ .`,
  `brm: Sat ~ cs(x1) + cs(x2)`,
  `ordinalNet: parallelTerms=FALSE, nonparallelTerms=TRUE`.

- `parallelism = ~ x1 + x2` (formula naming non-parallel terms):
  partial specification. Equivalent to `vglm: parallel = TRUE ~ -1 +
  (complement of formula)`, `clm: nominal = ~ x1 + x2`,
  `brm: Sat ~ Infl + cs(Cont)`,
  NOT SUPPORTED natively by `ordinalNet`.

- `parallelism = "semi"` (`ordinalNet`-specific): semi-parallel model with
  penalized nonparallel terms.

## Proposal 2: `clm`-style formula

Adopt `clm`'s convention (shared by `brm`): a formula naming the
**non-parallel** terms.

```
nominal = ~ x3
=> x3 is non-parallel; everything else is parallel.
```

Advantages:

- More intuitive for users who want to "break" the parallel regression
  assumption for specific variables.
- Maps to `ordinalNet`'s all-or-nothing flags: if all variables are in
  the formula, set `parallelTerms=FALSE, nonparallelTerms=TRUE`; if no
  variables are in the formula, set `parallelTerms=TRUE,
  nonparallelTerms=FALSE`. Partial parallelism is not supported natively
  (see Challenges below).
- `clm`, `brm`, and `ordinalNet` use the same direction (non-parallel terms).

### Implementation

Given the constraints, the most practical harmonization for ordered
would be to:

1. Register `nominal` (or a renamed `parallelism`) as a model argument
   for each engine, accepting a right-hand-side formula.

2. Document the translation:
   - For `clm`: pass directly as `nominal = ~ ...`.
   - For `vglm`: convert to `parallel = TRUE ~ -1 + (complement)`.
   - For `brm`: wrap non-parallel terms in `cs()` in the formula.
   - For `ordinalNet`: convert to `parallelTerms`/`nonparallelTerms`
     booleans, with an error if partial parallelism is requested.

3. Keep `parallelPenaltyFactor` as a separate engine-specific argument
   (`ordinalNet` only).

4. Document that `vglm`/`brm` and `clm` use opposite sign conventions for
   coefficients (cumulative probability vs. latent variable).

5. Note that `brm` is Bayesian; posterior means approximate MLEs with
   large samples but require MCMC computation time.

This approach respects each engine's native interface while giving users
clear guidance on how to express the same model across engines.

## Proposal 3: `cs()`-style formula syntax

The `gen_additive_mod()` model with the `vgam` engine demonstrates how
formula specials control model structure: the `s()` function marks
predictors for smooth treatment, and the formula is passed through to
`VGAM::vgam()` which interprets it natively. This precedent motivates a
similar approach for parallelism: a `cs()` ("category-specific") formula
special that marks predictors for non-parallel treatment.

Adopt `brm`'s `cs()` convention directly in the model formula (as
demonstrated in the `brm` section above): variables wrapped in `cs()` are
category-specific (non-parallel); unwrapped variables are parallel.
Unlike Proposal 2, this needs no separate argument --- the formula
itself carries the parallelism information.

```
Sat ~ Infl + cs(Cont)         # Infl parallel, Cont non-parallel
Sat ~ cs(Infl) + cs(Cont)     # All non-parallel
Sat ~ Infl + Cont             # All parallel
```

### Translation to `vglm`

The wrapper parses the formula to identify `cs()` terms, strips the
wrappers, and constructs the `parallel` argument for the family function.
`vglm` names the **parallel** terms, so the wrapper computes the complement
of the `cs()` variables:

```r
# User formula: Sat ~ Infl + cs(Cont)
# cs() vars: Cont  =>  parallel vars: Infl
# Translation:
vglm(
  Sat ~ Infl + Cont, data = data,
  family = VGAM::cumulative(link = "logitlink", parallel = TRUE ~ -1 + Infl)
)
```

When all terms are non-parallel (`cs(Infl) + cs(Cont)`), the wrapper
sets `parallel = FALSE`. When no terms are `cs()`, it sets
`parallel = TRUE`. The intercept is always category-specific regardless
of the specification.

### Translation to `clm`

The wrapper extracts `cs()` terms and passes them as the `nominal`
formula. This maps directly: `cs()` terms become nominal terms.

```r
# User formula: Sat ~ Infl + cs(Cont)
# cs() vars: Cont
# Translation:
clm(Sat ~ Infl + Cont, data = data, nominal = ~ Cont)
```

### Translation to `ordinalNet`

`ordinalNet` cannot express partial parallelism natively (see the
structural limitation discussed above). The wrapper checks whether the
`cs()` specification is all-or-nothing:

```r
# Sat ~ Infl + Cont (no cs()): fully parallel model
ordinalNet(x, y, ..., parallelTerms = TRUE, nonparallelTerms = FALSE)

# Sat ~ cs(Infl) + cs(Cont) (all cs()): fully non-parallel model
ordinalNet(x, y, ..., parallelTerms = FALSE, nonparallelTerms = TRUE)

# Sat ~ Infl + cs(Cont) (partial): ERROR
stop("ordinalNet does not support partial parallelism. ",
     "Use engine 'clm' or 'vglm' for partial parallelism.")
```

### Interaction with `s()` terms

`brm` allows `s()` and `cs()` in the same formula, but only on
**different** predictors:

```r
bf(rating ~ period + cs(treat) + s(x), family = sratio())
```

Here `period` and `x` have parallel effects and `treat` has a
non-parallel effect; `cs()` controls parallelism, `s()` controls
functional form. Nesting is not allowed --- `cs(s(x))` raises an error ---
and `cs()` works only with `sratio()`, `cratio()`, and `acat()`
families, not `cumulative()`.

`vglm` is richer here: smooth terms and non-parallel effects coexist
naturally, and a smooth term can itself be non-parallel, a capability
`brm` lacks. The catch is that `vglm`'s `parallel` formula matches terms by
**exact string label**, so a raw variable name like `x1` does not match
the smooth term label `s(x1, df = 2)`:

```r
vgam(y ~ s(x1, df = 2) + x2,
     family = VGAM::cumulative(link = "logitlink",
                         parallel = FALSE ~ 1 + s(x1, df = 2)))
```

A `cs()` implementation in ordered's vgam wrapper would therefore need
to map `cs()` variable names to their exact smooth term labels when
constructing the `parallel` formula.

### Implementation sketch

The wrapper would parse the formula to identify `cs()` specials,
analogous to how `VGAM::vgam()` handles `s()` specials:

```r
parse_cs_terms <- function(formula) {
  terms_obj <- terms(formula, specials = "cs")
  cs_idx <- attr(terms_obj, "specials")$cs
  all_vars <- rownames(attr(terms_obj, "factors"))
  if (is.null(cs_idx)) {
    return(list(formula = formula, cs_vars = character(0)))
  }
  cs_vars <- all_vars[cs_idx]
  clean_formula <- remove_cs_wrappers(formula)
  list(formula = clean_formula, cs_vars = cs_vars)
}
```

This approach is formula-native, matches the user's mental model ("wrap
in cs() to make that variable non-parallel"), and parallels the existing
`gen_additive_mod` + `s()` precedent in the tidymodels ecosystem.

## Challenges with harmonization

1. **Formula semantics differ**: In `vglm`, the formula names PARALLEL
   terms (`TRUE ~ -1 + x1` => x1 is parallel). In `clm` and `brm`, the
   formula names NON-PARALLEL terms (`nominal = ~ x1` or `cs(x1)`).
   These are logical inverses. A unified interface should pick ONE
   convention and translate for each engine.

2. **Family vs. model argument**: In `vglm`, the parallelism is specified
   inside the family function (`cumulative(parallel=...)`), not as a model
   argument. In `clm` and `ordinalNet`, it is a model argument. In `brm`, it
   is part of the formula (`cs()`). For a unified interface, it should be
   a model argument (consistent with `clm` and `ordinalNet`), and the wrapper
   should inject it into the `vglm` family constructor and the `brm`
   formula.

3. **Intercept handling**: `vglm` explicitly excludes the intercept from
   the parallelism constraint (intercepts are always category-specific).
   `clm` does the same by design. `ordinalNet` has no explicit intercept in
   the design matrix. A unified interface should document that intercepts
   are always category-specific regardless of the parallelism setting.

4. **Penalty interactions**: `ordinalNet`'s `parallelPenaltyFactor` has no
   analog in `vglm` or `clm`. If we harmonize the parallelism specification,
   we still need a separate mechanism for differential penalization. This
   could be a separate `parallel_penalty_factor` engine argument.

5. **`ordinalNet`'s all-or-nothing design**: `ordinalNet` cannot express
   partial parallelism natively. A formula-based interface would require
   `ordinalNet` to either accept the formula and zero out non-selected
   nonparallel terms, or raise an error when partial parallelism is
   requested. The first option is feasible but requires a wrapper that
   constructs the design matrix with interaction terms only for
   non-parallel variables.

6. **Threshold structure**: `clm` supports threshold constraints
   (equidistant, symmetric, etc.) that interact with nominal effects.
   `vglm` supports similar constraints via the `Thresh` argument. `brm`
   supports threshold constraints via the `threshold` argument in the
   family. `ordinalNet` has no threshold structure. A unified interface
   would need to decide whether threshold constraints are part of the
   parallelism specification or separate. See the "Threshold constraints
   with parallel regression" section above.

7. **Scale model**: `clm` supports a scale formula for heteroscedastic
   models. `vglm` supports this via `s()` in `vgam`. `brm` supports this
   via distributional regression (`bf(..., sigma ~ ...)`). `ordinalNet`
   does not. This is orthogonal to parallelism but adds complexity to the
   unified interface.

8. **Link function interactions**: `clm` supports Aranda-Ordaz and log-gamma
   links that `vglm`, `ordinalNet`, and `brm` do not. The parallelism
   specification is orthogonal to the link, but some links may have
   convergence issues with non-parallel models.
