---
title: "Fitting and Predicting with ordered"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Fitting and Predicting with ordered}
output:
  knitr:::html_vignette
---

```{r}
#| label: startup
#| include: false
library(utils)
```

These examples illustrate which models, engines, and prediction types
are available in ordered.
As a reminder, in parsnip,

- the **model type** differentiates basic modeling approaches,
  such as random forests and ordinal regression;

- the **mode** denotes in what kind of modeling context it will be used
  (here, classification); and

- the computational **engine** indicates how the model is fit,
  such as with a specific R package implementation
  or even methods outside of R like Keras or Stan.

The following examples use the same data set throughout.

## `decision_tree()` models

<details id="decision-tree-rpartScore">

<summary>With the `"rpartScore"` engine</summary>

We'll model satisfaction of householders under varying conditions.

```{r}
library(parsnip)
library(ordered)

house_data <-
  MASS::housing[rep(seq(nrow(MASS::housing)), MASS::housing$Freq), -5]
house_sample <- sort(sample(nrow(house_data), 12L))
house_train <- house_data[-house_sample, ]
house_test <- house_data[house_sample, ]
```

We can define the model with specific parameters:

```{r}
dt_spec <-
  decision_tree() |>
  set_engine("rpartScore") |>
  set_mode("classification")
dt_spec
```

Now we create the model fit object:

```{r}
set.seed(1)
dt_fit <- dt_spec |> fit(Sat ~ Infl + Type + Cont, data = house_train)
dt_fit
```

The holdout data can be predicted for the most likely class:

```{r}
predict(dt_fit, house_test, type = "class")
```

</details>

## `gen_additive_mod()` models

<details id="gen-additive-mod-vgam">

<summary>With the `"vgam"` engine</summary>

We'll model satisfaction of householders under varying conditions.

```{r}
library(parsnip)
library(ordered)

house_data <-
  MASS::housing[rep(seq(nrow(MASS::housing)), MASS::housing$Freq), -5]
house_sample <- sort(sample(nrow(house_data), 12L))
house_train <- house_data[-house_sample, ]
house_test <- house_data[house_sample, ]
```

We can define the model with specific parameters:

```{r}
gam_spec <-
  gen_additive_mod() |>
  set_engine("vgam") |>
  set_mode("classification")
gam_spec
```

Now we create the model fit object:

```{r}
set.seed(1)
gam_fit <- gam_spec |> fit(Sat ~ Infl + Type + Cont, data = house_train)
gam_fit
```

The holdout data can be predicted for the most likely class or all class probabilities:

```{r}
predict(gam_fit, house_test, type = "class")
predict(gam_fit, house_test, type = "prob")
```

</details>

## `ordinal_reg()` models

<details id="ordinal-reg-polr">

<summary>With the `"polr"` engine</summary>

We'll model satisfaction of householders under varying conditions.

```{r}
library(parsnip)
library(ordered)

house_data <-
  MASS::housing[rep(seq(nrow(MASS::housing)), MASS::housing$Freq), -5]
house_sample <- sort(sample(nrow(house_data), 12L))
house_train <- house_data[-house_sample, ]
house_test <- house_data[house_sample, ]
```

We can define the model with specific parameters:

```{r}
or_spec <-
  ordinal_reg(ordinal_link = "probit") |>
  set_engine("polr") |>
  set_mode("classification")
or_spec
```

Now we create the model fit object:

```{r}
set.seed(1)
or_fit <- or_spec |> fit(Sat ~ Infl + Type + Cont, data = house_train)
or_fit
```

The holdout data can be predicted for the most likely class or all class probabilities:

```{r}
predict(or_fit, house_test, type = "class")
predict(or_fit, house_test, type = "prob")
```

</details>

<details id="ordinal-reg-ordinalNet">

<summary>With the `"ordinalNet"` engine</summary>

We'll model satisfaction of householders under varying conditions.

```{r}
library(parsnip)
library(ordered)

house_data <-
  MASS::housing[rep(seq(nrow(MASS::housing)), MASS::housing$Freq), -5]
house_sample <- sort(sample(nrow(house_data), 12L))
house_train <- house_data[-house_sample, ]
house_test <- house_data[house_sample, ]
```

We can define the model with specific parameters:

```{r}
or_spec <-
  ordinal_reg(penalty = .001, mixture = .5) |>
  set_engine("ordinalNet") |>
  set_mode("classification")
or_spec
```

Now we create the model fit object:

```{r}
set.seed(1)
or_fit <- or_spec |> fit(Sat ~ Infl + Type + Cont, data = house_train)
or_fit
```

The holdout data can be predicted for the most likely class or all class probabilities:

```{r}
predict(or_fit, house_test, type = "class")
predict(or_fit, house_test, type = "prob")
```

</details>

<details id="ordinal-reg-vglm">

<summary>With the `"vglm"` engine</summary>

We'll model satisfaction of householders under varying conditions.

```{r}
library(parsnip)
library(ordered)

house_data <-
  MASS::housing[rep(seq(nrow(MASS::housing)), MASS::housing$Freq), -5]
house_sample <- sort(sample(nrow(house_data), 12L))
house_train <- house_data[-house_sample, ]
house_test <- house_data[house_sample, ]
```

We can define the model with specific parameters:

```{r}
or_spec <-
  ordinal_reg() |>
  set_engine("vglm") |>
  set_mode("classification")
or_spec
```

Now we create the model fit object:

```{r}
set.seed(1)
or_fit <- or_spec |> fit(Sat ~ Infl + Type + Cont, data = house_train)
or_fit
```

The holdout data can be predicted for the most likely class or all class probabilities:

```{r}
predict(or_fit, house_test, type = "class")
predict(or_fit, house_test, type = "prob")
```

</details>

## `rand_forest()` models

<details id="rand-forest-ordinalForest">

<summary>With the `"ordinalForest"` engine</summary>

We'll model satisfaction of householders under varying conditions.

```{r}
library(parsnip)
library(ordered)

house_data <-
  MASS::housing[rep(seq(nrow(MASS::housing)), MASS::housing$Freq), -5]
house_sample <- sort(sample(nrow(house_data), 12L))
house_train <- house_data[-house_sample, ]
house_test <- house_data[house_sample, ]
```

We can define the model with specific parameters:

```{r}
rf_spec <-
  rand_forest(trees = 1000) |>
  set_engine("ordinalForest") |>
  set_mode("classification")
rf_spec
```

Now we create the model fit object:

```{r}
set.seed(1)
rf_fit <- rf_spec |> fit(Sat ~ Infl + Type + Cont, data = house_train)
rf_fit
```

The holdout data can be predicted for the most likely class or all class probabilities:

```{r}
predict(rf_fit, house_test, type = "class")
predict(rf_fit, house_test, type = "prob")
```

</details>
